\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-09-30 paper.qmd
Added initial writeup for the supplementary aand acknowledgments

This study makes several important contributions to the field of NLDR. We have developed an algorithm to evaluate the most useful NLDR method and (hyper-)parameter choices for creating a reasonable \gD{} layout of high-dimensional data. Our objective is to fit a model for the \gD{} layout that preserves the relationships between neighboring points and turns it into a high-dimensional wireframe, which can be overlaid on the data and visualized using a tour. This approach is defined as *model-in-data-space*. Viewing a model in the data space is an ideal way to examine the fit.

The effectiveness of this approach is illustrated through various examples. For instance, the S-curve example demonstrates how the model accurately fits the points, capturing both local and global structures in high-dimensional space. Our simulation case study further, five Gaussian cluster example shows that while all observed NLDR methods preserve the global structure, only tSNE effectively maintains the local structure, highlighting the specific strengths and quirks of different methods.

Human behavior often shows a desire for more certainty and a tendency to prefer well-separated views. This emphasizes the importance of clear and distinct clusters. For example, in the UMAP layout of the **pbmc** dataset suggested by @chen2023, three distant, well-separated clusters are shown. However, our model reveals that these clusters are actually close to each other in \pD{}. Additionally, the model discovers non-uniform data distribution and non-linear structures within the clusters that are not visible in the UMAP layout, demonstrating the ability of our model in uncovering hidden data characteristics.

@@

Code, and data for reproducing this paper are available at [https://github.com/JayaniLakshika/paper-nldr-vis-algorithm](https://github.com/JayaniLakshika/paper-nldr-vis-algorithm).

>>>>

This study makes several important contributions to the field of NLDR. We have developed an algorithm to evaluate the most useful NLDR method and (hyper-)parameter choices for creating an accurate \gD{} layout of high-dimensional data. Our objective is to fit a model for the \gD{} layout that preserves the relationships between neighboring points and turns it into a high-dimensional wireframe, which can be overlaid on the data and visualized using a tour. This approach is defined as *model-in-data-space*. Viewing a model in the data space is an ideal way to examine the fit.

The effectiveness of this approach is illustrated through various examples. For instance, the two-curvy clusters example demonstrates how the model accurately fits the points, capturing both local and global structures in high-dimensional space. Our simulation case study further, five Gaussian cluster example shows that while all observed NLDR methods preserve the global structure, only tSNE effectively maintains the local structure, highlighting the specific strengths and quirks of different methods.

Human behavior often shows a desire for more certainty and a tendency to prefer well-separated views. This emphasizes the importance of clear and distinct clusters. For example, in the UMAP layout of the **PBMC3k** dataset suggested by @chen2023, three distant, well-separated clusters are shown. However, our model reveals that these clusters are actually close to each other in \pD{}. Additionally, the model discovers non-uniform data distribution and non-linear structures within the clusters that are not visible in the UMAP layout, demonstrating the ability of our model in uncovering hidden data characteristics.

@@

Appendix: The appendix includes more details about the hexagonal binning algorithm (appendix.pdf, Portable Document Format file).

R package `quollr`: The R package `quollr` containing codes to fit, and visualize the model. (need to add quollr .zip file, GNU zipped tar file)

## Acknowledgments

These `R` packages were used for the work:
The project's GitHub repository ([https://github.com/JayaniLakshika/paper-nldr-vis-algorithm](https://github.com/JayaniLakshika/paper-nldr-vis-algorithm)) contains all materials required to reproduce this article.

### 2024-09-30 paper.tex
Updated Figure 10
Added initial writeup for the supplementary aand acknowledgments

\caption{long edge removal}

@@

This study makes several important contributions to the field of NLDR.
We have developed an algorithm to evaluate the most useful NLDR method
and (hyper-)parameter choices for creating a reasonable \gD{} layout of
high-dimensional data. Our objective is to fit a model for the \gD{}
layout that preserves the relationships between neighboring points and
turns it into a high-dimensional wireframe, which can be overlaid on the
data and visualized using a tour. This approach is defined as
\emph{model-in-data-space}. Viewing a model in the data space is an
ideal way to examine the fit.

The effectiveness of this approach is illustrated through various
examples. For instance, the S-curve example demonstrates how the model
accurately fits the points, capturing both local and global structures
in high-dimensional space. Our simulation case study further, five
Gaussian cluster example shows that while all observed NLDR methods
preserve the global structure, only tSNE effectively maintains the local
structure, highlighting the specific strengths and quirks of different
methods.

Human behavior often shows a desire for more certainty and a tendency to
prefer well-separated views. This emphasizes the importance of clear and
distinct clusters. For example, in the UMAP layout of the \textbf{pbmc}
dataset suggested by \citet{chen2023}, three distant, well-separated
clusters are shown. However, our model reveals that these clusters are
actually close to each other in \pD{}. Additionally, the model discovers
non-uniform data distribution and non-linear structures within the
clusters that are not visible in the UMAP layout, demonstrating the
ability of our model in uncovering hidden data characteristics.

@@

Code, and data for reproducing this paper are available at
\url{https://github.com/JayaniLakshika/paper-nldr-vis-algorithm}.

>>>>

\caption{Wireframes with Delaunay triangulation, and two choices of edge
removal in \gD{} and \pD{}.}

@@

This study makes several important contributions to the field of NLDR.
We have developed an algorithm to evaluate the most useful NLDR method
and (hyper-)parameter choices for creating an accurate \gD{} layout of
high-dimensional data. Our objective is to fit a model for the \gD{}
layout that preserves the relationships between neighboring points and
turns it into a high-dimensional wireframe, which can be overlaid on the
data and visualized using a tour. This approach is defined as
\emph{model-in-data-space}. Viewing a model in the data space is an
ideal way to examine the fit.

The effectiveness of this approach is illustrated through various
examples. For instance, the two-curvy clusters example demonstrates how
the model accurately fits the points, capturing both local and global
structures in high-dimensional space. Our simulation case study further,
five Gaussian cluster example shows that while all observed NLDR methods
preserve the global structure, only tSNE effectively maintains the local
structure, highlighting the specific strengths and quirks of different
methods.

Human behavior often shows a desire for more certainty and a tendency to
prefer well-separated views. This emphasizes the importance of clear and
distinct clusters. For example, in the UMAP layout of the
\textbf{PBMC3k} dataset suggested by \citet{chen2023}, three distant,
well-separated clusters are shown. However, our model reveals that these
clusters are actually close to each other in \pD{}. Additionally, the
model discovers non-uniform data distribution and non-linear structures
within the clusters that are not visible in the UMAP layout,
demonstrating the ability of our model in uncovering hidden data
characteristics.

@@

Appendix: The appendix includes more details about the hexagonal binning
algorithm (appendix.pdf, Portable Document Format file).

R package \texttt{quollr}: The R package \texttt{quollr} containing
codes to fit, and visualize the model. (need to add quollr .zip file,
GNU zipped tar file)

\section{Acknowledgments}\label{acknowledgments}

These \texttt{R} packages were used for the work: The project's GitHub
repository
(\url{https://github.com/JayaniLakshika/paper-nldr-vis-algorithm})
contains all materials required to reproduce this article.
